Prefix caching reuses a transformer's prefill only across an exact shared prefix, so a single changed
field---a clock tick, a session id, an account-status flip---invalidates the entire downstream cache.
We begin from a puzzle: the cache region \emph{before} a field is provably reusable (key/value deviation
$0.0$), yet surgically overwriting only the field's own key/value vectors and reusing the rest leaves the
model acting on the \emph{old} value. We explain why, and turn the explanation into two capabilities.
Through four independent causal probes we show that at prefill the model computes the
\emph{field-conditioned conclusion} and stores it on downstream aggregator/delimiter tokens: the field's
own key/value drives less than $1\%$ of the decision, while the downstream ``notes'' drive essentially
all of it. The KV cache is thus a notebook of memoized conclusions, and two consequences follow.
\textbf{(i) It is editable:} because the conclusion is already written downstream, the right fix is not
recomputation but a salient \emph{erratum} that amends the notes; \texttt{field+erratum} matches the
strong hoist-to-end baseline without rewriting the prompt, and under chain-of-thought the cheap in-place
edit alone recovers the oracle decision ($0.94$ at $8$B) at ${\sim}1\%$ of the compute.
\textbf{(ii) It is composable:} the notes are localized, position-portable, and context-robust, so a
precompiled skill can be RoPE-repositioned and spliced into any context---behaviorally indistinguishable
from full recompute (logit cosine $0.90$--$0.999$ across twelve models) at $O(L)$ rather than $O(L^2)$
time-to-first-token ($13.9\times$ at $32$k tokens). A keystone experiment---editing a field \emph{inside}
a transplanted skill---reproduces the editing mechanism verbatim, showing edit and compose act on one
substrate; a unified agent over $10$ domains $\times$ $100$ trajectories is decision-identical to full
recompute (agreement $0.81$--$1.0$) at up to $14.9\times$ lower latency. The substrate is any per-token
attention KV cache: we validate it across scale, quantization, Mixture-of-Experts, multimodal image
caches, and---via small adapters---Multi-head Latent Attention and sliding-window and interleaved-M-RoPE
models, and we map exactly where it holds up to the 2026 sparse-attention frontier. The caching
machinery itself is prior art; our contributions are the mechanism, the unification, a decision-governance
evaluation, and the attention-variant adapters.
