\section{Application: editable and composable user memory}
\label{sec:memory}

A concrete, high-value instance of the edit+compose substrate is \emph{user memory}: the
large, dynamically-summarized profile of facts an assistant re-reads every turn. Memory is
big (thousands--tens of thousands of tokens), reused across turns, and \emph{mutated mid-session}
by tool calls---so where it lives in the prompt is a dilemma the mechanism of
\Cref{sec:mechanism} explains. Placed at the \emph{front} (\note{[sys][MEM][traj]}), the
trajectory memoizes memory-conditioned conclusions downstream, so a memory change forces a
costly downstream reprefill (and an in-place edit is ignored). Placed at the \emph{end}
(\note{[sys][traj][MEM]}), memory's KV depends on the whole trajectory and must be re-attended
every turn. We resolve this by treating memory as a \emph{skill that is also edited}:
precompile it once in isolation, place it late (\note{[sys][traj][MEM][query]} so the decode
reads it directly), RoPE-reposition it each turn, repair one boundary token, and edit it in
place when it changes.

\paragraph{Memory transplant is faithful, to 70B (E2).} Across ten models from $0.6$B to
$70$B, a precompiled+repositioned memory chunk reproduces the full-recompute decision logits
with cosine $0.94$--$0.9996$; a \emph{single} seam-repair token closes the start-of-chunk
boundary gap (Llama-3.1-8B late: cos $0.94\rightarrow0.994$). The cleanest decision-governance
test is Llama-3.1-70B, whose decisions \emph{genuinely vary} (so agreement is non-trivial,
unlike the constant-answer regime of smaller models): the late, seam-repaired transplant
reproduces the full-recompute decision $0.93$ of the time at logit cosine $0.997$, and
\emph{late placement beats early} ($0.93$ vs.\ $0.83$) exactly as the mechanism predicts
(the decode reads memory directly rather than through pre-digested notes across the transplant
boundary). The no-rotation control collapses (late decision agreement $0.18$ vs.\ $0.78$
rotated), confirming the re-rotation is necessary.

\paragraph{Memory is editable mid-session (E3).} When a stored fact is toggled, reusing the
\emph{stale} memory recovers the flipped decision essentially never ($\le0.03$). Every real
edit recovers it, and---consistent with \Cref{sec:editable}---under chain-of-thought the
\emph{near-free in-place edit} (one token recomputed) suffices, and \emph{strengthens with
scale}: correctness $0.92\rightarrow0.99\rightarrow1.00$ across Qwen3-1.7B/4B/14B (and $0.98$
at 32B). On models where the chain does not re-read the field, the append-only \erratum{} is
the robust fallback (McNemar \erratum{}$>$\inplace{} on Llama-3.1-8B, $p=0.031$). This is the editing axis
that concurrent KV-cache memory systems lack: MemArt \citep{memart2026} and EPIC
\citep{hu2025epic} retrieve and splice \emph{static} memory blocks position-independently but
have no in-place memory \emph{update}; our additions are the editing operation, the
decision-governance lens, and the mechanism (\Cref{sec:mechanism}) that explains why boundary
recompute suffices.

\paragraph{Keystone: editing inside transplanted memory reproduces the mechanism.} Editing a
field \emph{inside a transplanted memory chunk} in direct mode (Llama-3.1-70B, whose decisions
vary so recovery is measurable) reproduces \Cref{sec:mechanism}'s memoization verbatim:
refreshing the field's KV alone recovers only $0.55$ of the flipped decision (the conclusion was
memoized \emph{downstream}, not in the field), and recovery climbs monotonically as more
downstream tokens are recomputed ($0.55\rightarrow0.84$ at $K{=}16\rightarrow0.94$ full),
exactly as for a normally-prefilled context. Under chain-of-thought this stickiness
\emph{dissolves}---a selective@$K$ sweep is flat at ${\approx}0.98$ for all $K$ (the chain
re-reads the field, so $K^\star{\approx}1$). Edit and compose thus act on one substrate, and the
direct/CoT dissociation matches the editing law of \Cref{sec:editable}.

\paragraph{Real memory: LoCoMo external validity.} The synthetic gated decisions isolate the
mechanism; to test real memory we run the long-conversation QA benchmark LoCoMo
\citep{maharana2024locomo} (MemArt's setting): the multi-session dialogue (median
${\sim}19.7$k tokens) is the memory, precompiled and spliced before each question. Over
\emph{all} $1{,}540$ answerable questions per model, transplant is \emph{statistically
equivalent} to full recompute in QA accuracy on Qwen3-4B, 14B, and 32B (TOST, margin $0.03$:
$|\Delta|\le0.015$) and within a small $-2.7$ points on Llama-3.1-8B, with answer-token logit
cosine $0.991$--$0.998$ throughout. The \emph{compose} axis thus holds on
real conversational memory, not only synthetic decisions.

\paragraph{Granularity is a free knob (E4).} Splitting memory into $S$ independently-precompiled
blocks makes a localized edit $S\times$ cheaper (recompute one block) and is
\emph{decision-lossless} up to $S{=}16$ (agreement flat at $1.00$ on Qwen3-4B), because the
independent facts are integrated at read time, not within memory. Logit cosine degrades with $S$
($0.998\rightarrow0.953$) but \emph{identically whether the gating facts are contiguous in one
block or split across blocks}---splitting relevant facts across independently-precompiled blocks
does not specifically hurt; only genuinely cross-\emph{referential} facts (untested) should.

\paragraph{Placement and the pre-digestion cost (E1).} Pre-digestion is real, and we quantify
it. Under full recompute and chain-of-thought, reading memory \emph{late} carries a small but
\emph{statistically significant} accuracy cost vs.\ \emph{early} (Qwen3-4B, $N{=}192$:
GEE-logistic $\beta_{\text{late}}=-0.44$, $p=0.018$), and the cost \emph{grows with memory
length}: the early$-$late gap is ${\approx}0$ at $2$k tokens but ${+}0.09$ at $16$k and ${+}0.16$
at $32$k. The mechanism explains it---late placement forgoes prefill-time pre-digestion and
relies on the decode/CoT to integrate raw memory. This is a \emph{tradeoff, not free}: late
placement is still preferred because it enables $O(L)$ editing/transplant and $2.3$--$4.3\times$
TTFT (below) and the transplant is faithful at a fixed placement (above), so the net is small
accuracy for large efficiency---with early placement preferable when memory is very long and
accuracy-critical. (Direct one-shot decisions are at chance for the reasoning-native Qwen3 family
\emph{to 32B}, breaking only at Llama-3.1-70B, direct $0.81$; CoT is the operative regime across
scale---see appendix.)

\paragraph{End-to-end agent (E5).} We implement a live agent (\Cref{app:memory-agent}) that
composes memory once, re-rotates it each turn, and edits it on tool-driven changes. Over
$24$ sessions $\times$ $12$ turns with $\approx$2k-token memories, against a
reprefill-every-turn-at-the-end baseline it cuts cumulative time-to-first-token by
$2.3$--$4.32\times$ (growing with model size, to 32B), and against front-placement
reprefill-on-change by $1.4$--$3.27\times$, while reproducing the full-reprefill next-token logits at a
token-matched oracle (cosine $0.98$--$0.99$). One honest caveat: greedy chains-of-thought are
sensitive to sub-percent logit differences, so the exact reasoning \emph{chain} is not always
reproduced (chain agreement $0.31$--$0.69$) even though the next-token decision is faithful
and CoT \emph{accuracy} is comparable; the clean decision-governance equivalence is the
short-context editing result above (\Cref{sec:editable}-style, agreement $0.89$--$0.95$). User
memory is thus a second instance of one substrate that is both editable and composable.
