\section{Related work}
\label{sec:related}

\paragraph{Where computation is stored, and how to edit it.} A line of interpretability work localizes
stored \emph{knowledge} in transformer weights and edits it there: ROME and MEMIT \citep{meng2022rome,
meng2023memit} locate and rewrite factual associations in MLP weights, while circuit analyses such as the
indirect-object-identification study \citep{wang2023ioi} trace how specific computations are carried by
attention heads. Weight editing targets \emph{durable, global} facts; we compare against a faithful ROME
and a LoRA fine-tune empirically (\Cref{tab:weight}) and find them ill-suited to \emph{mutable per-request}
state---a global edit contaminates concurrent requests and damages unrelated decisions---which is exactly
the niche the editable KV cache fills. Closest in spirit, \citet{lindsey2025biology} find models commit \emph{plans} to
specific tokens (e.g.\ a planned rhyme stored on a line-break token) during the forward pass. We study a
complementary object: not weights and not decode-time computation, but the \emph{KV cache}---the
activations an inference system already stores and an editor can directly manipulate---and show it holds
\emph{memoized conclusions} concentrated on aggregator/delimiter tokens. To our knowledge this is the
first causal account of why in-place KV field editing fails and what to do instead.

\paragraph{KV reuse and composable caching.} Reusing precomputed KV beyond an exact prefix is an active
systems topic, and our \emph{composing} capability builds directly on it; we claim none of the caching
machinery as novel. Prompt Cache \citep{gim2024promptcache} precomputes reusable prompt modules with
position placeholders and splices them. CacheBlend \citep{yao2025cacheblend} reuses non-prefix chunk KV
and selectively recomputes ${\sim}15\%$ of tokens to restore cross-attention. EPIC
\citep{hu2025epic} introduces position-independent caching with \textsc{AttnLink}, recomputing only a few
chunk-boundary tokens (exploiting attention sinks) for near-linear recompute. CacheSlide
\citep{liu2026cacheslide} reuses KV in a position-\emph{aware} way via relative-position-dependent caching,
and MPIC \citep{mpic2025} extends position-independent caching to the multimodal setting by recomputing
image-boundary tokens; KVLink \citep{yang2025kvlink} is a further reuse system. Mapped onto our terms, our
RoPE-repositioning is CacheSlide's relative-position reuse, our seam-repair is the boundary recompute of
CacheBlend/EPIC/MPIC, and our image-KV transplant is MPIC's idea (we \emph{re-rotate} M-RoPE rather than
recompute). Our contributions over this line are orthogonal: (i) the \emph{mechanism} that explains
\emph{why} boundary recompute is what is needed; (ii) a \emph{decision-governance} evaluation---does a
transplanted skill still \emph{govern the tool decision}, rather than only preserve perplexity or
throughput; (iii) the \emph{editing} axis and the keystone unification; and (iv) adapters that extend the
operations to new attention representations (MLA, interleaved M-RoPE, sliding-window).

\paragraph{Prefix caching, KV compression, and reuse systems.} Production prefix caching (vLLM Automatic
Prefix Caching, SGLang RadixAttention) reuses exact prefixes. A large literature instead \emph{compresses}
or \emph{evicts} the cache: StreamingLLM keeps attention sinks \citep{xiao2024streamingllm}; H2O
\citep{zhang2023h2o}, Scissorhands \citep{liu2023scissorhands}, and SnapKV \citep{li2024snapkv} evict
low-importance tokens; Quest \citep{tang2024quest} keeps all tokens but attends sparsely. Serving systems
stream or share cached KV across requests---CacheGen \citep{liu2024cachegen} for fast loading and RAGCache
\citep{jin2024ragcache} for retrieval reuse. These change \emph{which} tokens are present, and compose with
our edit/transplant operations only over retained tokens; we treat them, the latent/decoupled-RoPE
representation of Multi-head Latent Attention \citep{deepseekv2,deepseekv3}, and the 2026
sparse/compressed-attention designs as scope boundaries in \Cref{sec:reach}. Our repositioning relies on
rotary position embeddings \citep{su2024rope} and their extensions \citep{peng2024yarn}.

\paragraph{Activation-level interventions.} Beyond weight editing, a body of work intervenes on
\emph{activations}: steering and inference-time interventions \citep{li2023iti}, task and function vectors
\citep{ilharco2023taskvectors,todd2024functionvectors}, and the causal-mediation/activation-patching
methodology we adapt \citep{vig2020causal}. These edit residual-stream directions to change behavior; we
instead read and write the \emph{KV cache} itself---the per-token activations a serving system already
persists---which is what makes the intervention both interpretable and deployable.

\paragraph{Agents and tool use.} Our evaluation targets tool-using agents in the style of ReAct
\citep{yao2023react} and Toolformer \citep{schick2023toolformer}, and we measure end-to-end task success on
the $\tau^2$-bench tool-agent-user benchmark \citep{barres2025tau2}, the dual-control successor of
$\tau$-bench \citep{yao2024taubench}, where state changes mid-trajectory make
cache staleness a first-class correctness problem.
