\section{Conclusion}
\label{sec:conclusion}

A surgical edit to a field's KV is ignored not because the cache is fragile but because the model has
already done the work: at prefill it computes the field-conditioned \emph{conclusion} and writes it onto
downstream aggregator tokens, so the decision only reads those notes back. This reframes what a KV cache
is---not a write-once byproduct of prefill, but a structured record of intermediate conclusions that we can
read, amend, and rearrange. Two capabilities follow from the one mechanism: \emph{editing} the notes (a late,
salient erratum in place of recomputation) and \emph{composing} them (repositioning and splicing a
precompiled skill in $O(L)$ time); a keystone experiment shows both act on the same notes. The account holds
wherever attention keeps per-token KV, up to the block-granular frontier of 2026 sparse-attention designs
\citep{deepseekv32,deepseekv4}.

The question this opens is larger than either operation. If prefill routinely deposits reusable conclusions
into the cache, then much of what a model has worked out mid-context is already written down somewhere we can
inspect and overwrite. Editing
and composition are two uses of that record, and they work today even though no model was trained for them.
The larger opportunity is to make this explicit: models trained to be \emph{aware} that their KV can be
composed and edited---exposing notes that are cleanly addressable, splice-able, and revisable by design---would
turn the cache from a linear, append-only log into a \emph{programmable} memory the system reads, writes, and
rearranges. Once the KV cache is programmable rather than merely linear, context engineering changes
shape significantly: \emph{skills, memory, and state become first-class, reusable cache objects} rather than text re-prefilled
on every turn. We offer \emph{models take notes at prefill} as a first step toward that programmable KV
cache---a notebook the model keeps for itself, and one it can learn to read and edit.
